```{r}
# libraries to import

library(arrow)
library(tidyverse)
library(ggplot2)

```

```{r}

data <- read_parquet("trots_2013-2022.parquet")
head(data, 50)

summary(data$Saddlecloth)

unique(data$Saddlecloth)

# convert the saddle cloth data into integer

generate_twenty <- function(lst) {
  for (i in 1:20) {
    lst[i] <- as.character(i)
  }
  return(lst)
}

saddle_levels <- character(20)


data$Saddlecloth <- factor(data$Saddlecloth, levels=generate_twenty(saddle_levels))

unique(data$Saddlecloth)

data$FinishPosition <- str_trim(data$FinishPosition)
unique(data$FinishPosition)

#data %>% select(GoingAbbrev, RaceID, GoingID, ClassRestriction)
unique(data$ClassRestriction)
unique(data$GoingAbbrev)
unique(data$GoingID)
```

```{r}

winning_data <- data %>% filter(FinishPosition == "1" | FinishPosition == "2" | FinishPosition == "3" | FinishPosition == "4" | FinishPosition == "5" | FinishPosition == "6" | FinishPosition == "7")

head(winning_data$FinishPosition, 20)

winning_data$FinishPosition <- as.numeric(winning_data$FinishPosition)

head(winning_data$FinishPosition, 20)

```

```{r}

ggplot(data=winning_data, mapping=aes(x=Saddlecloth, y=FinishPosition)) +
  geom_boxplot()
```

```{r}
mdl <- lm(data=winning_data, FinishPosition ~ Saddlecloth)

summary(mdl)

# Conclusion 1: there are minor differences depending on the choice of the saddle cloth (i.e. Saddlecloth = 1 results in the estimation of 4.2 FinishPosition while Saddlecloth = 18 results in the estimation of 3.8 finish position). This doesn't seem like a significant predictor for deciding the winning probability.
```

```{r}

losing_data <- data %>% filter(FinishPosition != "1", FinishPosition != "2", FinishPosition != "3", FinishPosition != "4", FinishPosition != "5", FinishPosition != "6", FinishPosition != "7")


nrow(losing_data)
unique(losing_data$FinishPosition)

losing_data_category <- losing_data %>% filter(FinishPosition == "BS" | FinishPosition == "UN" | FinishPosition == "DQ" | FinishPosition == "PU" | FinishPosition == "NP" | FinishPosition == "FL" | FinishPosition == "UR" | FinishPosition == "WC") %>% mutate(FinishPosition = "50")

losing_data_numeric <- losing_data %>% filter(FinishPosition != "1", FinishPosition != "2", FinishPosition != "3", FinishPosition != "4", FinishPosition != "5", FinishPosition != "6", FinishPosition != "7", FinishPosition != "BS", FinishPosition != "UN", FinishPosition != "DQ", FinishPosition != "PU", FinishPosition != "NP", FinishPosition != "FL", FinishPosition != "UR", FinishPosition != "WC")

unique(losing_data_category$FinishPosition)

unique(losing_data_numeric$FinishPosition)

losing_data <- rbind(losing_data_category, losing_data_numeric)

nrow(losing_data)
```

```{r}
losing_data$FinishPosition <- as.numeric(losing_data$FinishPosition)

finishpos_saddlecloth <- ggplot(data=losing_data, aes(x=Saddlecloth, y=FinishPosition)) + 
  geom_boxplot()

finishpos_saddlecloth

# Conclusion 2: Saddle cloth type seems to greatly affect the finish position result. We should include saddle cloth as one of our predictors. There seems to be a trend where certain type of saddlecloth results in lower chance of being disqualified by mutating losing_data_category to 50. 
```

```{r}

mdl_two <- lm(data=losing_data, FinishPosition ~ Saddlecloth)

summary(mdl_two)

# Conclusion: Saddle cloth type seems to greatly affect the finish position result. We should include saddle cloth as one of our predictors.

# I subdivided finish position into two: "winning" (rank 1 to 7) and "not winning" (rank 8 to 18 + disqualified horses). Note: all disqualified horses were grouped into a single rank as 50 to give higher weight towards the rank. From rank 1-7, there were no significant difference in wearing different saddle cloths. From not winning horses, I noticed that saddle cloth 14, 15, 16, 17 all had significantly lower chance of being disqualified. I have done my analysis by giving the weight of being disqualified as 50. I started to see significant difference in rank (for instance rank goes down from 33 to 37 from one saddle cloth to another).
```

# SexRestriction -> Analysis on C&G group competing in other competition

```{r}

# Check if there's a lot of C&G in the entire dataset

test_1 <- data %>% filter(SexRestriction == "C&G")
nrow(test_1)

# We notice there's 120745 C&G data

# Check if the age of C&G group exceeds age 4

test_2 <- data %>% filter(SexRestriction == "C&G") %>% filter(HorseAge >= 5)
nrow(test_2)

# Interestingly, female horses were also be in category C&G 

data %>% filter(SexRestriction == "C&G") %>% filter(Gender == "F")

# I've decided to assume that C&G group is for young male horses (age below 4) that have been castrated. Interestingly, there were some female horses that were part of C&G category. I decided to ignore these female horses that were part of C&G category as well because there were not many data with this condition. 

# Recall: C for male horses age less than 4 and G for male horses that have been castrated. C&G is not restricted to horses of C AND G; I figured there's a case where it was C OR G but this was also a very rare. With this, I'll just assume that C&G is looking at C AND G (i.e, age below 4 and castrated). 
```

```{r}

data %>% filter(RaceID==1621386) %>% select(FinishPosition, RaceID, Gender, SexRestriction, HorseAge)

data %>% filter(RaceID==1584675) %>% select(FinishPosition, RaceID, Gender, SexRestriction, HorseAge)

data %>% filter(RaceID==1596840) %>% select(FinishPosition, RaceID, Gender, SexRestriction, HorseAge)

data %>% filter(RaceID==1621385) %>% select(FinishPosition, RaceID, Gender, SexRestriction, HorseAge)

```

# Additional investigation: For a competition including both genders, does male horses have significantly higher chance in winning the race compared to female horses?

```{r}
female_vs_male <- data %>% filter(SexRestriction == "") %>% filter(FinishPosition == "1" | FinishPosition == "2" | FinishPosition == "3" | FinishPosition == "4" | FinishPosition == "5" | FinishPosition == "6" | FinishPosition == "7"  | FinishPosition == "8") %>% select(SexRestriction, Gender, FinishPosition)
```

```{r}
both_sex <- data %>% filter(SexRestriction == "") %>% select(SexRestriction, Gender, FinishPosition, RaceID) %>% group_by(RaceID) %>% summarise(both=length(unique(Gender)))

#nrow(both_sex)
#nrow(both_sex %>% filter(Gender=="M"))
```

```{r}
both_sex_races <- both_sex %>% filter(both == 2)

nrow(both_sex_races)

test_both_sex_races <- data %>% filter(RaceID %in% both_sex_races$RaceID) %>% select(SexRestriction, Gender, Prizemoney, RaceID, Disqualified, FinishPosition, AgeRestriction, HorseAge)
```

```{r}

test_data_two <- test_both_sex_races %>% group_by(RaceID) %>% summarise(count = n()) %>% filter(count < 8)

```

```{r}

test_data_three <- test_both_sex_races %>% filter(Gender == "F") %>% filter(FinishPosition == "1" | FinishPosition == "2" | FinishPosition == "3") %>% select(HorseAge, AgeRestriction, FinishPosition, Gender, RaceID)

test_data_four <- test_both_sex_races %>% filter(Gender == "M") %>% filter(FinishPosition == "1" | FinishPosition == "2" | FinishPosition == "3") %>% select(HorseAge, AgeRestriction, FinishPosition, Gender, RaceID)

```


```{r}
test_both_sex_races %>% filter(RaceID == 1598497) %>% select(SexRestriction, Gender, Prizemoney, RaceID, Disqualified, FinishPosition, AgeRestriction, HorseAge)

```

```{r}

summary(lm(data=test_both_sex_races, Prizemoney ~ Gender))

#nrow(test_data %>% filter(Gender == "M"))

#nrow(test_data %>% filter(Gender == "F"))

nrow(test_both_sex_races %>% filter(Disqualified == FALSE) %>% filter(Gender == "M") %>% filter(FinishPosition == "1" | FinishPosition == "2" | FinishPosition == "3" |FinishPosition == "4" |FinishPosition == "5" | FinishPosition == "6" | FinishPosition == "7")) / nrow(test_data %>% filter(Gender == "M"))

nrow(test_both_sex_races %>% filter(Disqualified == FALSE) %>% filter(Gender == "F") %>% filter(FinishPosition == "1" | FinishPosition == "2" | FinishPosition == "3" |FinishPosition == "4" |FinishPosition == "5" | FinishPosition == "6" | FinishPosition == "7")) / nrow(test_data %>% filter(Gender == "F"))


```

# StartingLine -> maybe associated with handicap distance further study needed

```{r}
handicap <- data$HandicapDistance
head(handicap)
unique(handicap)
# handicap distances are: -50, -25, 0, 25, 50, 75 

starting <- data$StartingLine
head(starting)
unique(starting)
# starting line: 1, 2, -1

# Question: is starting line 1 => handicap distance of X? 
# This hypothesis was not true. There's no direct relationship.
  
nov <- data %>% filter(StartingLine == 1) %>% select(HandicapDistance)

nrow(nov)

# Starting line = 1 means handicap distance = 0?

dec <- data %>% filter(StartingLine == -1) %>% select(HandicapDistance)

dec

# Starting line = -1, 1M, rest 10k, 6k

#There's no direct relationship between starting line and handicap distance. 
#In a way, those with handicap had to stay in starting line -1. For those who did not have any handicap distance stayed in either line 1 or 2. Most data were with starting line of -1. 

jan <- data %>% filter(StartingLine == -1) %>% filter(StartType == "V") %>% select(StartType)

jan

# Starting Line == -1, 1, 2 -> starttype == M
# Starting Line == -1 -> mostly starttype == V
# then is startline 1 vs 2 => both M

negative_start_line <- data %>% filter(StartingLine == -1) %>% filter(StartType == "M") %>% select(RaceID, StartingLine, FinishPosition, StartType)

data %>% filter(RaceID %in% negative_start_line$RaceID) %>% filter(StartingLine != -1) %>% select(RaceID, StartingLine, FinishPosition, StartType)

#data %>% filter(RaceID == 1641580) %>% select(RaceID, StartingLine, FinishPosition, StartType, HandicapDistance)

# It's possible to have StartingLine = -1, 1, 2 altogether in a single race. However, this is extremely rare (3 races) and there is at most ONE horse that starts with StartingLine = -1. Thus, we can safely assume that for StartType == "M", StartingLine is either (1 and 2) or ONLY -1.
```

# Is starting line 1 better compared to starting line 2?

```{r}

# Trying this with winning_data 

# start_one in winning_data

start_one <- winning_data %>% filter(StartingLine == 1) %>% select(FinishPosition, StartingLine)

start_two <- winning_data %>% filter(StartingLine == 2) %>% select(FinishPosition, StartingLine)

# starting line: 1, 2, -1

nrow(start_one) / nrow(data %>% filter(StartingLine == 1) %>%  select(FinishPosition, StartingLine))

nrow(start_two) / nrow(data %>% filter(StartingLine == 2) %>%  select(FinishPosition, StartingLine))
```

```{r}

# compared to start_two

start_two <- winning_data %>% filter(StartingLine == 2) %>% select(FinishPosition, StartingLine)
nrow(start_two)

# We see that for starting line == 1, there's 62786 data that belongs to Finish Position 1 to 7 whereas for line == 2, there's 28300 data that belongs to Finish Position that is 8-18 and disqualified. This intuitively seems like starting line 1 has more benefit towards a chance of winning for the case of rank 1 to 7. 

# start_one in losing_data

start_one_losing <- losing_data %>% filter(StartingLine == 1) %>% select(FinishPosition, StartingLine)
start_one_losing

nrow(start_one_losing)

start_two_losing <- losing_data %>% filter(StartingLine == 2) %>% select(FinishPosition, StartingLine)
start_two_losing
nrow(start_two_losing)

# For losing_data it seems like 43571 cases were starting line == 1 whereas 40897 cases were at starting line = 2. I'm not sure if this is supposed to imply something significant

# Just by looking at the result, it seems like for the Finish Position 1 to 7, starting line 1 has advantage over 2 significantly (about 2 times). For Finish position that did not lead to winning, it seems like there also exist a similar trend where starting line 1 is more advantageous compared to starting line 2. However, the difference is quite small. Overall, we can conclude that there's starting line = 1 is more advantageous compared to starting line = 2. 
```

# NoFrontCover -> needs further analysis especially for value -9

```{r}

unique(data$NoFrontCover)
# how often does -9 show up in our data? is -9 supposed to mean a missing data? => there's quite a lot of -9 showing up in our entire data so we can't ignore this -9

# based on my research it seems like usually in horse racing, Good cover is particularly valuable in longer races where conserving energy is essential (having a cover is most likely better than not having a cover)

# but what is -9? => then what if it was just a short distance competition so they decided not to write 1 or 0 since it's meaningless? 

# or since we can't figure what this -9 is, can we look into whether this is related to the winning rate?
```

```{r}

# NoFrontcover = -9 > NoFrontcover = 0 > NoFrontcover = 1 for number of cases

temp_race_data <- data %>% filter(NoFrontCover == 0) %>% filter(PositionInRunning == 18) %>% select(NoFrontCover, WideOffRail, PositionInRunning, RaceID, Distance, FinishPosition)

temp_race_data

# When NoFrontCover == 0, PIR can never be 1

```
# TASK #1 WideOffRail <- find if this matters for winning probability
```{r}
wideoff <- data$WideOffRail
# ggplot(data=data, aes(x=wideoff, y=FinishPosition)) + geom_boxplot()

# or can we look at it this way?
unique(data$WideOffRail)
# -9  1  2  3  5  4  6

wideoff_one <- data %>% filter(WideOffRail == 1) %>% filter(FinishPosition == 1| FinishPosition == 2| FinishPosition == 3| FinishPosition == 4| FinishPosition == 5|FinishPosition == 6| FinishPosition == 7) %>% select(WideOffRail, FinishPosition)

wideoff_one_prob <- nrow(wideoff_one)/nrow(data %>% filter(WideOffRail == 1 | WideOffRail == 2 | WideOffRail == 3| WideOffRail == 4| WideOffRail == 5| WideOffRail == 6)) 

wideoff_two <- data %>% filter(WideOffRail == 2) %>% filter(FinishPosition == 1| FinishPosition == 2| FinishPosition == 3| FinishPosition == 4| FinishPosition == 5|FinishPosition == 6| FinishPosition == 7) %>% select(WideOffRail, FinishPosition)

wideoff_two_prob <- nrow(wideoff_two)/nrow(data %>% filter(WideOffRail == 1 | WideOffRail == 2 | WideOffRail == 3| WideOffRail == 4| WideOffRail == 5| WideOffRail == 6)) 

wideoff_three <- data %>% filter(WideOffRail == 3) %>% filter(FinishPosition == 1| FinishPosition == 2| FinishPosition == 3| FinishPosition == 4| FinishPosition == 5|FinishPosition == 6| FinishPosition == 7) %>% select(WideOffRail, FinishPosition)

wideoff_three_prob <- nrow(wideoff_three)/nrow(data %>% filter(WideOffRail == 1 | WideOffRail == 2 | WideOffRail == 3| WideOffRail == 4| WideOffRail == 5| WideOffRail == 6)) 
wideoff_four <- data %>% filter(WideOffRail == 4) %>% filter(FinishPosition == 1| FinishPosition == 2| FinishPosition == 3| FinishPosition == 4| FinishPosition == 5|FinishPosition == 6| FinishPosition == 7) %>% select(WideOffRail, FinishPosition)

wideoff_four_prob <- nrow(wideoff_four)/nrow(data %>% filter(WideOffRail == 1 | WideOffRail == 2 | WideOffRail == 3| WideOffRail == 4| WideOffRail == 5| WideOffRail == 6))

wideoff_five <- data %>% filter(WideOffRail == 5) %>% filter(FinishPosition == 1| FinishPosition == 2| FinishPosition == 3| FinishPosition == 4| FinishPosition == 5|FinishPosition == 6| FinishPosition == 7) %>% select(WideOffRail, FinishPosition)

wideoff_five_prob <- nrow(wideoff_five)/nrow(data %>% filter(WideOffRail == 1 | WideOffRail == 2 | WideOffRail == 3| WideOffRail == 4| WideOffRail == 5| WideOffRail == 6))

wideoff_six <- data %>% filter(WideOffRail == 6) %>% filter(FinishPosition == 1| FinishPosition == 2| FinishPosition == 3| FinishPosition == 4| FinishPosition == 5|FinishPosition == 6| FinishPosition == 7) %>% select(WideOffRail, FinishPosition)

wideoff_six_prob <- nrow(wideoff_six)/nrow(data %>% filter(WideOffRail == 1 | WideOffRail == 2 | WideOffRail == 3| WideOffRail == 4| WideOffRail == 5| WideOffRail == 6)) 

wideoff_neg_nine <- data %>% filter(WideOffRail == -9) %>% filter(FinishPosition == 1| FinishPosition == 2| FinishPosition == 3| FinishPosition == 4| FinishPosition == 5|FinishPosition == 6| FinishPosition == 7) %>% select(WideOffRail, FinishPosition)

wideoff_neg_nine_prob <- nrow(wideoff_neg_nine)/nrow(data %>% filter(WideOffRail == 1 | WideOffRail == 2 | WideOffRail == 3| WideOffRail == 4| WideOffRail == 5| WideOffRail == 6)) 
wideoff_one_prob
wideoff_two_prob
wideoff_three_prob
wideoff_four_prob
wideoff_five_prob
wideoff_six_prob
wideoff_neg_nine_prob
```
# TASK 2

# NoFrontCover = 1 vs NoFrontCover = 0, find if this matters for winning probability

```{r}

nofront <- data %>% filter(NoFrontCover == 1) %>% filter(FinishPosition == 1| FinishPosition == 2| FinishPosition == 3| FinishPosition == 4| FinishPosition == 5| FinishPosition == 6| FinishPosition == 7) %>% select(NoFrontCover, FinishPosition)

nrow(nofront)/nrow(data %>% filter(NoFrontCover == 1| NoFrontCover == 0| NoFrontCover == -9)%>% filter(FinishPosition == 1| FinishPosition == 2| FinishPosition == 3| FinishPosition == 4| FinishPosition == 5| FinishPosition == 6| FinishPosition == 7))

yesfront <- data %>% filter(NoFrontCover == 0) %>% filter(FinishPosition == 1| FinishPosition == 2| FinishPosition == 3| FinishPosition == 4| FinishPosition == 5| FinishPosition == 6| FinishPosition == 7) %>% select(NoFrontCover, FinishPosition) 

nrow(yesfront)/nrow(data %>% filter(NoFrontCover == 1| NoFrontCover == 0| NoFrontCover == -9)%>% filter(FinishPosition == 1| FinishPosition == 2| FinishPosition == 3| FinishPosition == 4| FinishPosition == 5| FinishPosition == 6| FinishPosition == 7))

nofront_negnine <- data %>% filter(NoFrontCover == -9) %>% filter(FinishPosition == 1| FinishPosition == 2| FinishPosition == 3| FinishPosition == 4| FinishPosition == 5| FinishPosition == 6| FinishPosition == 7) %>% select(NoFrontCover, FinishPosition)

nrow(nofront_negnine)/nrow(data %>% filter(NoFrontCover == 1| NoFrontCover == 0| NoFrontCover == -9)%>% filter(FinishPosition == 1| FinishPosition == 2| FinishPosition == 3| FinishPosition == 4| FinishPosition == 5| FinishPosition == 6| FinishPosition == 7))


```

# Position in running = -9, WideOffRail = -9 are always together with NoFrontCover = -9

```{r}
no_front_neg_nine <- data %>% filter(NoFrontCover == -9) %>% select(NoFrontCover)
nrow(no_front_neg_nine)

# there's 778645 cases where it's written as NoFrontCover = -9

no_front_one <- data %>% filter(NoFrontCover == 1) %>% select(NoFrontCover)
nrow(no_front_one)

no_front_zero <- data %>% filter(NoFrontCover == 0) %>% select(NoFrontCover)
nrow(no_front_zero)
nrow(data)

neg_nine <- data %>% filter(NoFrontCover == -9)
neg_nine

# I found out that if NoFrontCover == -9,  then PositionInRunning == -9 and WideOffRail == -9. 

unique(data$WideOffRail)
# for WideOffRail: -9, 1, 2, 3, 4, 5, 6

# NOTE: if the shape of the track is oval/circular it seems like wideoffrail matters whereas if it was just straight line track, the wideoffrail doesn't matter. what if this neg 9 is for the cases where it was straight line track so there's no need to talk about it and instead wrote -9? => based on the analysis we figured this isn't true
```



